Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-11-15 src/tex/ms.tex
Sect. -> Section, Eqn. -> Equation

\section{Baseline Model}\label{sec:met_baseline}
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of a habitable zone inner edge discontinuity (see diagram in Figure~\ref{fig:flowchart}).
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, and the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

@@

\subsection{Transit probability}
We model the occurrence of transits by assuming isotropic orientations of planetary orbits and calculating the impact parameters $b = a \cos(i)/R_\star$.
Following the approach in \citet{Bixel2021}, we further consider only planets with $|b| < 1$.
For these cases we calculate the transit depth
\begin{equation}\label{eq:transitdepth}
    \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2,
\end{equation}
which is relevant for the detection probability of the respective planet~(see Sect.~\ref{sec:sensitivity}).
Excluding all non-transiting planets diminishes the sample to \var{percentage_transiting} of its original size.

@@

\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Sect.~\ref{sec:syn_planets}).
This case serves as our null hypothesis.

@@

A sensible choice of priors is central for evidence estimation via nested sampling.
As the parameters of interest are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
For $S_\mathrm{thresh}$, we chose a uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
We sampled $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Sect.~\ref{sec:met_rghmodel}).
For $f_\mathrm{rgh}$, we chose a uniform prior in $[0, 1]$.
Finally, we adopted a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
In the case of measuring bulk densities instead of transit radii, we drew uniformly from $[1, 6] \, \SI{}{\gram\per\centi\meter\cubed}$.

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
    \script{optimistic_RS.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\subsubsection{Additional planet mass measurements}\label{sec:res_followup}
Comparing a measurement of the habitable zone inner edge discontinuity in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.

@@

\subsubsection{Constraining the threshold instellation}\label{sec:res_constrain-S_thresh}
\begin{figure}[ht!]
    \script{plato_Sthresh_grid.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/S_thresh_posteriors.pdf}
        \caption{
            Retrieved posterior distributions of the threshold instellation for different survey realizations.
            All cases assume $x_{H_2O} = \var{wrr}$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
            Orange lines show the true value of the injected signal.
            Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy and precision.
        }
        \label{fig:posterior_surveys}
    \end{centering}
\end{figure}
\rev{Although the exact location of the runaway greenhouse transition depends on the stellar energy actually absorbed by the planet, }a measurement of $S_\mathrm{thresh}$ \rev{would be a} key constraint resulting from a detection \rev{of the habitable zone inner edge discontinuity}.
Here, we assess the ability of different mission concepts explored above to constrain this parameter.
Figure~\ref{fig:posterior_surveys} shows posterior distributions from a grid of inferences together with the true values of the injected signal for different fractions $f_\mathrm{rgh}$.
We consider three cases: only radius measurements, radius and mass measurements, and radius and mass measurements of only planets orbiting M~dwarfs.
The simulations otherwise represent the simulated \plato\ example described above.
We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Sect.~\ref{sec:res_samplesize}.

@@

\subsection{Statistical imprint of exoplanet climates}
We showed in Sect.~\ref{sec:res_signature} that injecting the theoretically predicted radius inflation effect into a synthetic planet population following the currently known demographics leaves a distinct pattern in the radius and density distribution as a function of instellation.
The transit radius inflation was suggested before as an observational diagnostic to probe the runaway greenhouse transition~\citep{Turbet2019}, and our simulations show quantitatively how the contributions of the individual radius changes combine to create a significant demographic signal.
Its strength depends on largely unknown factors such as the planetary volatile content, but turned out to be well in the detectable range under reasonable assumptions of these factors.

@@

\subsection{Detectability of the habitable zone inner edge discontinuity}
We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
Of course, even assuming capable instrumentation and an optimal survey design, our current models may not capture fully the complexity of trends in the geophysics and demographics of small planets.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and the diverse outcomes of rocky planet formation and evolution.
In the following, we will explore the key factors that may influence the emerging demographic signal.

@@

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
%discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
%\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

@@

Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
Similar science objectives are pursued by the \life\ initiative~\citep[][]{Quanz2022}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
However, because mid-infrared retrievals feature reduced degeneracy between cloud albedo and changes in surface area, the planet radius can be constrained in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
Since the peak thermal emission in runaway greenhouse atmospheres will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

@@

\subsubsection{The value of follow-up campaigns}
The constraining power changes when additional information beyond planet radii is available for the characterized planet population.
As runaway greenhouse phases leave a stronger imprint on bulk density than on planet radius (see Sect.~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
For a mission design similar to \plato\, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
At a fixed sample size, key parameters of the runaway greenhouse models can be more narrowly constrained when additional mass measurements are available.

@@

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume- or magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs is \rev{the extended runaway greenhouse phases of their planets that can last on} the order of gigayears~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

\subsection{Constraining planetary habitability}\label{sec:habitability}
%\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1. commonly used as environmental marker to assess habitability, and 2. influences the planet's climate (greenhouse effect).}
A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its surface habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
Its detection would thus not only empirically confirm the habitable zone concept but also help to locate it in the observationally available planetary parameter space.
In Sect.~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
A mission like \plato\ is well equipped to perform this measurement; the constraining power is directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.

@@

Once a runaway greenhouse region is identified in the parameter space, the community will  have a tool at hand to discern potentially habitable planets from Venusian worlds on an empirical basis.
Together with atmospheric measurements (see Sect.~\ref{sec:dis_atmospheres}), we will be able to put a number on the probability of an individual planet to harbor sufficient surface water to sustain life.

@@

\subsubsection{Bulk water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions\rev{~\citep{Selsis2007,Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}}.
Our nominal case assumes a bulk water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity. %, and the hypothesis tests are only marginally sensitive to $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

@@

\appendix
\section{Robustness tests}
\subsection{Alternative statistics for the average radius or bulk density}\label{app:binnedstats}
The hypothesis tests introduced in Sect.~\ref{sec:met_surveys-hypotests} rely on a statistical estimator for the variation of planetary radii or bulk densities as a function of net instellation, and we chose a moving average for this estimator in our nominal setup.
Here, we explore how robust our results are against this choice by demonstrating the recovery of the runaway greenhouse signal in the case of the optimistic survey (Sect.~\ref{sec:res_testability}) with an alternative estimator: instead of computing moving averages, we used a binned statistic.

@@

\begin{figure}[ht!]
    \script{optimistic_RS_binned.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
        \caption{
            Detection of the runaway greenhouse threshold with a binned statistic.
            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis. Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space. Random draws from the posterior of the runaway greenhouse hypothesis (Eqn.~\ref{eq:rgh_hypo}) are shown in red.
            As in the nominal case, the pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S_binned}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
The accuracy of the recovered instellation threshold is comparable.
This test demonstrates that our results are not sensitive to the choice of statistical estimator to test the hypotheses against.

@@

\subsection{Statistical imprint of runaway greenhouse atmospheres}\label{app:model-pop_comparison}
The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and \rev{bulk} water mass fraction~(compare Sect.~\ref{sec:met_rghmodel}).
In order to better understand the interaction between the planetary populations underlying our simulations and these predictions, we compared the latter to the average radius and bulk density changes we measured in the synthetic population.
We used the ``optimistic'' scenario with a sample size of \var{N_optimistic}.

>>>>

\section{Baseline Model}\label{sec:met_baseline}
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of a habitable zone inner edge discontinuity (see diagram in Figure~\ref{fig:flowchart}).
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Section~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, and the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

@@

\subsection{Transit probability}
We model the occurrence of transits by assuming isotropic orientations of planetary orbits and calculating the impact parameters $b = a \cos(i)/R_\star$.
Following the approach in \citet{Bixel2021}, we further consider only planets with $|b| < 1$.
For these cases we calculate the transit depth
\begin{equation}\label{eq:transitdepth}
    \delta = \left( \frac{R_\mathrm{P}}{R_\star} \right)^2,
\end{equation}
which is relevant for the detection probability of the respective planet~(see Section~\ref{sec:sensitivity}).
Excluding all non-transiting planets diminishes the sample to \var{percentage_transiting} of its original size.

@@

\textit{Non-runaway} planets retain the radius assigned based on exoplanet occurrence rates (see Section~\ref{sec:syn_planets}).
This case serves as our null hypothesis.

@@

A sensible choice of priors is central for evidence estimation via nested sampling.
As the parameters of interest are poorly constrained by previous data, we used relatively uninformative priors to sample the entire physically plausible parameter space.
For $S_\mathrm{thresh}$, we chose a uniform prior in $[\prSmin,\prSmax]\, \SI{}{\watt\per\meter\squared}$.
We sampled $x_{H_2O}$ from a log-uniform distribution to imply scale-invariant ignorance.
Its boundaries $[\prWRRmin, \prWRRmax]$ are motivated by the water mass fractions covered by the geophysical models~(Section~\ref{sec:met_rghmodel}).
For $f_\mathrm{rgh}$, we chose a uniform prior in $[0, 1]$.
Finally, we adopted a broad, uniform prior for $\langle R_\mathrm{P}\rangle_\mathrm{out}$ bound by $[\prRmin, \prRmax] \, \mathrm{R_\oplus}$.
In the case of measuring bulk densities instead of transit radii, we drew uniformly from $[1, 6] \, \SI{}{\gram\per\centi\meter\cubed}$.

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
    \script{optimistic_RS.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Equation~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is $f_\mathrm{rgh} = \frgh$, and we chose a water mass fraction of $x_{H_2O} = \var{wrr}$ for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\subsubsection{Additional planet mass measurements}\label{sec:res_followup}
Comparing a measurement of the habitable zone inner edge discontinuity in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
With an optimistic choice of geophysical parameters (see Section~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.

@@

\subsubsection{Constraining the threshold instellation}\label{sec:res_constrain-S_thresh}
\begin{figure}[ht!]
    \script{plato_Sthresh_grid.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/S_thresh_posteriors.pdf}
        \caption{
            Retrieved posterior distributions of the threshold instellation for different survey realizations.
            All cases assume $x_{H_2O} = \var{wrr}$ and a planet sample size $N=100\pm10$; The fraction of planets with runaway greenhouse climates varies across rows.
            Orange lines show the true value of the injected signal.
            Accuracy and precision of the constraint on $S_\mathrm{thresh}$ generally improve with higher $f_\mathrm{rgh}$.
            Bulk density-based inferences improve the constraints, and M~dwarf samples yield the highest accuracy and precision.
        }
        \label{fig:posterior_surveys}
    \end{centering}
\end{figure}
\rev{Although the exact location of the runaway greenhouse transition depends on the stellar energy actually absorbed by the planet, }a measurement of $S_\mathrm{thresh}$ \rev{would be a} key constraint resulting from a detection \rev{of the habitable zone inner edge discontinuity}.
Here, we assess the ability of different mission concepts explored above to constrain this parameter.
Figure~\ref{fig:posterior_surveys} shows posterior distributions from a grid of inferences together with the true values of the injected signal for different fractions $f_\mathrm{rgh}$.
We consider three cases: only radius measurements, radius and mass measurements, and radius and mass measurements of only planets orbiting M~dwarfs.
The simulations otherwise represent the simulated \plato\ example described above.
We chose a planetary sample size of $N=100\pm10$, as this was found to be a threshold case in Section~\ref{sec:res_samplesize}.

@@

\subsection{Statistical imprint of exoplanet climates}
We showed in Section~\ref{sec:res_signature} that injecting the theoretically predicted radius inflation effect into a synthetic planet population following the currently known demographics leaves a distinct pattern in the radius and density distribution as a function of instellation.
The transit radius inflation was suggested before as an observational diagnostic to probe the runaway greenhouse transition~\citep{Turbet2019}, and our simulations show quantitatively how the contributions of the individual radius changes combine to create a significant demographic signal.
Its strength depends on largely unknown factors such as the planetary volatile content, but turned out to be well in the detectable range under reasonable assumptions of these factors.

@@

\subsection{Detectability of the habitable zone inner edge discontinuity}
We showed in Section~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
Of course, even assuming capable instrumentation and an optimal survey design, our current models may not capture fully the complexity of trends in the geophysics and demographics of small planets.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and the diverse outcomes of rocky planet formation and evolution.
In the following, we will explore the key factors that may influence the emerging demographic signal.

@@

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
%discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
%\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Section 4]{Turbet2019}.}

@@

Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
Similar science objectives are pursued by the \life\ initiative~\citep[][]{Quanz2022}, a mission concept utilizing a space-based mid-infrared nulling interferometer.
Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
However, because mid-infrared retrievals feature reduced degeneracy between cloud albedo and changes in surface area, the planet radius can be constrained in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
Since the peak thermal emission in runaway greenhouse atmospheres will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Section~\ref{sec:results_FGK_M}).
With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

@@

\subsubsection{The value of follow-up campaigns}
The constraining power changes when additional information beyond planet radii is available for the characterized planet population.
As runaway greenhouse phases leave a stronger imprint on bulk density than on planet radius (see Section~\ref{sec:res_followup}), it would be beneficial to obtain constraints on planetary masses and test the runaway greenhouse hypothesis in density space instead of radius space.
This way, useful results can be obtained under more pessimistic conditions, e.g., a low predominant water content of planetary surfaces and atmospheres or a smaller available planet sample.
For a mission design similar to \plato\, a density-based hypothesis test on about a third of the overall sample is equivalent to a pure radius-based analysis.
At a fixed sample size, key parameters of the runaway greenhouse models can be more narrowly constrained when additional mass measurements are available.

@@

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Section~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume- or magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs is \rev{the extended runaway greenhouse phases of their planets that can last on} the order of gigayears~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

\subsection{Constraining planetary habitability}\label{sec:habitability}
%\todo[inline]{Magma oceans influence the amount of water available at the surface and atmosphere, which is 1. commonly used as environmental marker to assess habitability, and 2. influences the planet's climate (greenhouse effect).}
A potential for liquid water on the surface of a planet is commonly used as an environmental marker to assess its surface habitability~\citep{Huang1959,Hart1978,Kasting1993,Kaltenegger2011,Kopparapu2013}.
The runaway greenhouse transition represents an upper bound on received irradiation for this condition.
Its detection would thus not only empirically confirm the habitable zone concept but also help to locate it in the observationally available planetary parameter space.
In Section~\ref{sec:res_constrain-S_thresh}, we show that the threshold instellation at which the runaway greenhouse transition occurs can be reasonably constrained without imposing overly optimistic conditions on the underlying planet population, instrumentation, or survey strategy.
A mission like \plato\ is well equipped to perform this measurement; the constraining power is directly proportional to the proportion of characterized planets around M dwarfs and to the number of planets for which masses can be determined.

@@

Once a runaway greenhouse region is identified in the parameter space, the community will  have a tool at hand to discern potentially habitable planets from Venusian worlds on an empirical basis.
Together with atmospheric measurements (see Section~\ref{sec:dis_atmospheres}), we will be able to put a number on the probability of an individual planet to harbor sufficient surface water to sustain life.

@@

\subsubsection{Bulk water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions\rev{~\citep{Selsis2007,Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}}.
Our nominal case assumes a bulk water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity. %, and the hypothesis tests are only marginally sensitive to $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Section~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

@@

\appendix
\section{Robustness tests}
\subsection{Alternative statistics for the average radius or bulk density}\label{app:binnedstats}
The hypothesis tests introduced in Section~\ref{sec:met_surveys-hypotests} rely on a statistical estimator for the variation of planetary radii or bulk densities as a function of net instellation, and we chose a moving average for this estimator in our nominal setup.
Here, we explore how robust our results are against this choice by demonstrating the recovery of the runaway greenhouse signal in the case of the optimistic survey (Section~\ref{sec:res_testability}) with an alternative estimator: instead of computing moving averages, we used a binned statistic.

@@

\begin{figure}[ht!]
    \script{optimistic_RS_binned.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S_binned.pdf}
        \caption{
            Detection of the runaway greenhouse threshold with a binned statistic.
            Using simulated data from the optimistic survey case ($N = \var{N_optimistic}$, compare Figure~\ref{fig:optimistic_R-S}), we tested the runaway greenhouse hypothesis. Instead of using a running mean, we computed a binned statistic (arithmetic mean and standard deviation in blue) to assign planets the average radius or bulk density in their neighborhood in instellation space. Random draws from the posterior of the runaway greenhouse hypothesis (Equation~\ref{eq:rgh_hypo}) are shown in red.
            As in the nominal case, the pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S_binned}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S_binned} shows the simulated data together with binned, average planet radii and draws from the posterior of the hypothesis test.
A clear detection resulted, although with somewhat lower significance ($\Delta \ln \mathcal{Z} \approx 30$) compared to the nominal setup.
The accuracy of the recovered instellation threshold is comparable.
This test demonstrates that our results are not sensitive to the choice of statistical estimator to test the hypotheses against.

@@

\subsection{Statistical imprint of runaway greenhouse atmospheres}\label{app:model-pop_comparison}
The predicted runaway greenhouse-induced planet radius changes are a function of instellation, planet mass, and \rev{bulk} water mass fraction~(compare Section~\ref{sec:met_rghmodel}).
In order to better understand the interaction between the planetary populations underlying our simulations and these predictions, we compared the latter to the average radius and bulk density changes we measured in the synthetic population.
We used the ``optimistic'' scenario with a sample size of \var{N_optimistic}.
